### 2024-12-30 sections/2-methods.tex
wip: changes by NI

\begin{equation}
    \frac{\text{log} \left( \frac{|B_X(q, r_1)|}{|B_X(q, r_2)|} \right) }{\text{log} \left( \frac{r_1}{r_2} \right) }
    \label{eq:methods:lfd-original}
\end{equation}
where $B_X(q, r)$ is the set of points contained in the metric ball of radius $r$ centered at a point $q$ in the dataset $\textbf{X}$.
Intuitively, LFD measures the rate of change in the number of points in a ball of radius $r$ around a point $q$ as $r$ increases. When the vast majority of points in the dataset have low ($\ll D$) LFD, we can simply say that the dataset has low LFD.
We stress that this concept differs from the \textit{embedding dimension} of a dataset.
To illustrate the difference, consider the SILVA 18S rRNA dataset which contains genomic sequences with unaligned lengths of up to 3,718 base pairs and aligned length of 50,000 base pairs.
Hence, the \textit{embedding dimension} of this dataset is at least 3,718 and at most 50,000.
However, physical constraints (namely, biological evolution and biochemistry) constrain the data to a lower-dimensional manifold within this space.
LFD is an approximation of the dimensionality of that lower-dimensional manifold in the ``vicinity'' of a given point.
Figure~\ref{fig:results:lfd-plots} illustrates this concept on a variety of datasets, showing how real datasets uphold the manifold hypothesis.

@@

We define a \textit{cluster} as a set of points with a \textit{center} and a \textit{radius}.
The center is the geometric median of the points (or a smaller sample of the points) in the cluster, and so it is a real data point.
The radius is the maximum distance from the center to any point in the cluster.
Each non-leaf cluster has two child clusters in much the same way that a node in a binary tree has two child nodes.
Note that clusters can have overlapping volumes and, in such cases, points in the overlapping volume are assigned to exactly one of the overlapping clusters.
As a consequence, a cluster can be a proper subset of the metric ball at the same center and radius, i.e. $C(c, r) \subset B_X(c, r)$.

Hereafter, when we refer to the LFD of a cluster, it is estimated at the length scale of the cluster radius and half that radius.
This allows us to rewrite Definition~\ref{eq:methods:lfd-original} as
$
    \log_2 \left( \frac{|C(c, r)|}{|C(c, \frac{r}{2})|} \right)
    \label{eq:methods:lfd-simplified}
$
where $|C(c, r)|$ is the cardinality of the cluster $C$ with center $c$ and radius $r$, and $C(c, \frac{r}{2})$ is the set of points in $C$ which are no more than a distance $\frac{r}{2}$ from $c$, i.e. $C(c, \frac{r}{2}) = \{p : p \in C \land f(c, p) \leq \frac{r}{2}\}$.

The \textit{metric entropy} $\mathcal{N}_{r}(X)$ for some radius $r$ was defined, in~\cite{yu2015entropy}, as the minimum number of clusters of a uniform radius $r$ needed to cover the data.
In this paper, where the clustering is hierarchical rather than flat, we define the metric entropy $\mathcal{N}_{\hat{r}}(X)$ of a dataset $X$ as the number of leaf clusters in the tree where $\hat{r}$ is the mean radius of all leaf clusters.

We start by performing a divisive hierarchical clustering on the dataset using CLAM.
The procedure is similar to that outlined in CHESS~\cite{ishaq2019clustered}, but with
better selection of poles for partitioning (see Algorithm~\ref{alg:methods:partition}) and depth-first reordering of the dataset (see Section~\ref{sec:methods:clustering:depth-first-reordering}).

@@

CLAM builds a binary tree of clusters using a divisive hierarchical clustering algorithm.
For a cluster $C$ with $|C|$ points, we begin by taking a random sample of $\sqrt{|C|}$ of its points, and computing pairwise distances between all points in this sample.
Using these distances, we compute the \textit{geometric median} of this sample; in other words, we find the point which minimizes the sum of distances to all other points in the sample.
This geometric median is the center of $C$.

The \textit{radius} of $C$ is the maximum distance from the \textit{center} to any other point in $C$.
The point which is responsible for that radius (i.e., the furthest point from the center) is designated the \textit{left pole} and the point which is furthest from the left pole is designated the \textit{right pole}.
We then partition the cluster into a \textit{left child} and a \textit{right child}, where the left child contains all points in the cluster which are closer to the left pole than to the right pole, and the right child contains all points in the cluster which are closer to the right pole than to the left pole.
Without loss of generality, we assign to the left child those points which are equidistant from the two poles.
Starting from a root cluster containing the entire dataset, we repeat this procedure until each leaf contains only one datum, or we meet some other user-specified stopping criteria, for example, minimum cluster radius, minimum cluster cardinality, maximum tree depth, etc.
This process is described in Algorithm \ref{alg:methods:partition}.
During the partitioning process, we also compute (and cache) the LFD of each cluster using Equation~\ref{eq:methods:lfd-simplified}.

\begin{algorithm} % enter the algorithm environment
    \caption{Partition($C$, $criteria$)} % give the algorithm a caption
    \label{alg:methods:partition} % and a label for \ref{} commands later in the document
    \begin{algorithmic}[0] % enter the algorithmic environment
        \Require $C$, a cluster
        \Require $criteria$, user-specified stopping criteria

@@

% % WE WILL USE THIS FOR COMPRESSION. DO NOT DELETE

% \subsubsection{Scaling Behavior of Cluster Radii}
% \label{sec:methods:clustering:scaling-behabior-of-cluster-radii}

% While it may be tempting to assume that cluster radii decrease with each application of Partition (refer to Algorithm \ref{alg:methods:partition}), unfortunately, this assumption is incorrect.
% Fortunately, we \textit{can} make some guarantees about the scaling behavior of cluster radii;
% in particular, we prove in this section that cluster radii will stop increasing after at most $d$ partitions, where $d$ is the fractal dimensionality of the dataset.

% We can describe a $d$-dimensional distribution of data by choosing some set of $d$ mutually orthogonal axes.
% Let $2R$ be the maximum pairwise distance among the points in the dataset.
% We choose the axes such that the two points that are $2R$ apart lie \textit{along one of the axes}.
% Thus, a $d$-dimensional hyper-sphere of radius $R$ would bound the dataset.
% In the worst case, (i.e., with a uniform-density distribution that fills the $d$-sphere), our axes will be such that $2R$ is the maximum pairwise distance \textit{along every axis}.
% Such a distribution would also produce a balanced cluster tree.

% In this case, Partition will select a maximally distant pair of points to use as poles;
% without loss of generality, say it chooses two points at the extrema one of the $d$ axes.
% After one partition, the maximum pairwise distance along that axis will be halved, i.e. bounded above by $R$.
% The next recursive Partition will select another of the $d$ axes, and once again, the distance along that axis will be bounded above by $R$
% Thus, after at most $d$ partitions, the maximum pairwise distance along each axis will be bounded above by $R$.
% The overall (i.e., not restricted to one axis) maximum pairwise distance
% will be bounded above by $R\sqrt{2}$ by, for example, two instances that lie at the extrema of different axes.
% See Figure~\ref{fig:methods:radii-scaling-behavior} for an example.

% Thus, starting with a cluster $C$ of radius $R$, after at most $d$ Partitions, the descendants of $C$ will each have radius
% bounded above by $\frac{R}{\sqrt{2}}$. In other words, cluster radii are guaranteed to decrease by a multiplicative factor of $\frac{\sqrt{2}}{2}$ after at
% most $d$ partitions.

% Note that, in practice, we never see a balanced clustering.
% Partition produces unbalanced trees due to the varying density of the sampling in different regions of the manifold and the low dimensional ``shape'' of the manifold.
% Further, in practice, the cluster radii decrease by a factor much larger than $\frac{\sqrt{2}}{2}$ with almost every partition;
% the upper bound of $d$ partitions is seldom realized.

% \begin{figure}[ht!]
%     \centering
%     \includegraphics[width=3.4in]{images/geometry/geometry.pdf}
%     \caption{
%         Scaling behavior of radii on a uniform-density disk, which represents the worst-case scenario for a two-dimensional distribution.
%         $R_0$ denotes the radius of the root cluster $C_0$, and $o_0$ denotes its center.
%         After one application of Algorithm~\ref{alg:methods:clustering:partition}, we have $C_1$, with radius $R_1$ and center $o_1$.
%         The right triangle formed by $o_0$, $o_1$, and $y_+$ in $C_1$ shows that $R_0 < R_1$.
%         Hence, the radius of a child cluster can be larger than its parent.
%         However, after another application of Algorithm~\ref{alg:methods:clustering:partition}, we have consumed all $d$ orthogonal axes,
%         as shown in $C_2$.
%         Now, clearly $R_2 < R_0$.
%         In fact, for each subsequent application of Algorithm~\ref{alg:methods:clustering:partition}, the radius of the resulting cluster is bounded above by $\frac{\sqrt{2}R}{2}$.
%     }
%     \label{fig:methods:radii-scaling-behavior}
% \end{figure}

@@

\subsection{The Search Problem}
\label{sec:methods:the-search-problem}

Given Algorithm~\ref{alg:methods:partition}, we can now formally pose the $k$-NN and $\rho$-NN search problems.

Given a query $q$, along with a distance function $f$ defined on a dataset $\textbf{X}$, we define the $k$-NN search problem, which aims to find the $k$ closest points to $q$ in $\textbf{X}$;
in other words, $k$-NN search aims to find the set $H$ such that $|H| = k$ and $H = B_X(q, \rho_k)$ where $\rho_k = \max \left\{ f(q, p) \ \forall \ p \in H \right\}$ is the distance from $q$ to the $k^{th}$ nearest neighbor in $\textbf{X}$.
We also have the $\rho$-NN search problem, which aims to find the all points in $\textbf{X}$ that are no more than a distance $\rho$ from $q$: the set $H = B_X(q, \rho) = \{x \in \textbf{X}: f(q, x) \leq \rho \}$.

Given a cluster $C$, let $c$ be its center and $r$ be its radius.
Our $\rho$-NN and $k$-NN search algorithms make use of the following properties, as illustrated in Figure~\ref{fig:methods:deltas}.

\begin{itemize}
    \item $\delta = f(q, c)$, the distance from the query to the cluster center $c$.
    \item $\delta^{+} = \delta + r$, the distance from the query to the theoretically farthest point in $C$.
    \item $\delta^{-} = \text{max}(0, \delta - r)$, the distance from the query to the theoretically closest point in $C$.
\end{itemize}

% \begin{figure}[ht!]
%     \centering
%     \includegraphics[scale=0.3]{images/geometry/deltas.pdf}
%     \caption{{\color{blue}$\delta$}, {\color{red}$\delta^{+}$}, and {\color{green}$\delta^{-}$} for a cluster $C$ and a query $q$.}
%     \label{fig:methods:deltas}
% \end{figure}

We define a \textit{singleton} as a cluster which either contains a single point (i.e., has cardinality 1) or which contains only duplicates of the same point.
A singleton clearly has zero radius, and so $\delta = \delta^{-} = \delta^{+}$.
Hence, we overload the above notation to also refer to the distance from a query to an individual point.

\subsection{\texorpdfstring{$\rho$}{p}-Nearest Neighbors Search}
\label{sec:methods:rnn-search}

We conduct $\rho$-NN search as described in~\cite{ishaq2019clustered}, but with the following improvement:
when a cluster overlaps with the query ball, instead of always proceeding to search both of its children, we proceed only with those children which might contain points in the query ball.

\begin{figure}[ht!]
    \begin{subfigure}{.5\textwidth}
        \centering
        \includegraphics[scale=0.4]{images/geometry/deltas.pdf}
        \caption{{\color{blue}$\delta$}, {\color{red}$\delta^{+}$}, and {\color{green}$\delta^{-}$} for a cluster $C$ and a query $q$.}
        \label{fig:methods:deltas}
      \end{subfigure}%
      \begin{subfigure}{.5\textwidth}
        \centering
        \includegraphics[scale=0.4]{images/geometry/overlapping-children-3.pdf}
        \caption{The geometry of a query ball overlapping with a cluster and either one or both of its children. Here, $l$ is the left pole, $r$ is the right pole, and $q$ is the query. Other points and distances are described in the text.}
        \label{fig:methods:overlapping-children}
      \end{subfigure}%
    \caption{CAKES uses geometric properties of clusters.}
\end{figure}

To determine whether both children can contain points in the query ball, we consider Figure~\ref{fig:methods:overlapping-children}.
Here, we overload the notation for $\overline{x y}$ to refer both to the line segment joining points $x$ and $y$ as well as to the length of that line segment.

Let $q$ denote the query, $\rho$ denote the search radius, and $l$ and $r$ denote the cluster's left and right poles respectively (see Section~\ref{sec:methods:clustering:building-the-tree}).
Without loss of generality, we assume that $\overline{q r \vphantom{l}} \leq \overline{q l}$.
Now let $q'$ be the projection of $q$ onto $\overline{l r}$, $m$ be the midpoint of $\overline{l r}$, and $d$ be the distance from $q'$ to $m$.
As a consequence of how we assign a point in the parent cluster to the left child in Algorithm~\ref{alg:methods:partition}, if $\rho < d$, then the left child cannot contain points inside the query ball.
In such a case we proceed to search only the right child.
Otherwise, we proceed with both children.

To check whether $d \leq \rho$, we note that $d = \overline{m q' \vphantom{l}} = \overline{m r \vphantom{l}} - \overline{q' r \vphantom{l}} = \frac{\overline{l r}}{2} - \overline{q' r \vphantom{l}}$.
Let $\theta$ denote $\angle l r q$, as shown in Figure~\ref{fig:methods:overlapping-children}.
By the Law of Cosines on $\triangle l r q$, we have that $\text{cos}(\theta) = \tfrac{\overline{l r}^2 + \ \overline{q r \vphantom{l}}^2 - \ \overline{q l}^2}{2 \cdot \overline{l r} \cdot \overline{q r \vphantom{l}}}$.
Since $\triangle r q q'$ is a right triangle, we also have that $\text{cos}(\theta) = \tfrac{\overline{q' r \vphantom{l}}}{\overline{q r \vphantom{l}}}$.
Combining the previous two equations and solving for $\overline{q' r \vphantom{l}}$, we have that $\overline{q' r \vphantom{l}} = \tfrac{\overline{q r \vphantom{l}}^2 + \ \overline{l r}^2 - \ \overline{q l}^2}{2 \cdot \overline{l r}}$.
Substituting for $\overline{q' r \vphantom{l}}$ in the equation for $d$, we have that $d = \tfrac{\overline{l r}}{2} - \tfrac{\overline{q r \vphantom{l}}^2 + \ \overline{l r}^2 - \ \overline{q l}^2}{2 \cdot \overline{l r}} = \tfrac{\overline{q l}^2 - \overline{q r \vphantom{l}}^2}{2 \cdot \overline{l r}}$.

Thus, $d \leq \rho \iff (\overline{q l} + \overline{q r \vphantom{l}})(\overline{q l} - \overline{q r \vphantom{l}}) \leq 2 \cdot \overline{l r} \cdot \rho$. Note, in particular, that this only requires distances between real points, and so it can be used with any distance function, even when $q'$ and $m$ are not real points or cannot be imputed from the data.

To perform $\rho$-NN search, we first perform a coarse \textit{tree-search}, as outlined in Algorithm~\ref{alg:methods:rnn-search:tree-search}, to find the leaf clusters which overlap with the query ball or any clusters which lie entirely within the query ball.
Then, for all such clusters, we perform a finer-grained \textit{leaf-search}, as outlined in Algorithm~\ref{alg:methods:rnn-search:leaf-search}, to find all points which are no more than a distance $\rho$ from the query.
The asymptotic complexity of $\rho$-NN is the same as in~\cite{ishaq2019clustered} and shown in Equation~\ref{eq:methods:rnn-search-complexity}.

\begin{minipage}{0.6\textwidth}
\begin{algorithm}[H]
    \caption{tree-search($C$, $q$, $\rho$)}
    \label{alg:methods:rnn-search:tree-search}
    \begin{tiny}
    \begin{algorithmic}[0]
        \Require $C$, a cluster
        \Require $q$, a query
        \Require $\rho$, a search radius

        \If{$C$ is a leaf \textbf{or} $\delta^+_C \leq \rho$}
            \State \textbf{return} $\{C\}$
        \Else

            \State $[l, r]$ $\Leftarrow$ \textit{pole}s of $C$
            \State $[L, R]$ $\Leftarrow$ \textit{children} of $C$

            \If{$\overline{q l} < \overline{q r \vphantom{l}}$}
                \State $[r, l] \Leftarrow [l, r]$
                \State $[R, L] \Leftarrow [L, R]$
            \EndIf

            \If{$(\overline{q l} + \overline{q r \vphantom{l}})(\overline{q l} - \overline{q r \vphantom{l}}) \leq 2 \cdot \overline{l r} \cdot \rho$}
                \State \textbf{return} tree-search($L, q, \rho$) $\cup$ tree-search($R, q, \rho$)
            \Else
                \State \textbf{return} tree-search($R, q, \rho$)
            \EndIf
        \EndIf
    \end{algorithmic}
    \end{tiny}
\end{algorithm}
\end{minipage}
\hfill
\begin{minipage}{0.35\textwidth}
\begin{algorithm}[H]
    \caption{leaf-search($S$, $q$, $\rho$)}
    \label{alg:methods:rnn-search:leaf-search}
    \begin{tiny}
    \begin{algorithmic}[0]
        \Require $S$, a set of clusters
        \Require $q$, a query
        \Require $\rho$, a search radius

        \State $H \Leftarrow \emptyset$, a set of hits

        \For{$C \in S$}
            \If{$\delta^+_C \leq \rho$}
                \State $H$ $\Leftarrow$ $H \cup \{C\}$
            \Else
                \For{$p \in C$}
                    \If{$f(p, q) \leq \rho$}
                        \State $H$ $\Leftarrow$ $H \cup \{p\}$
                    \EndIf
                \EndFor
            \EndIf
        \EndFor

        \State \textbf{return} $H$
    \end{algorithmic}
    \end{tiny}
\end{algorithm}
\end{minipage}

\begin{algorithm}
    \caption{$\rho$-NN-search($root$, $q$, $\rho$)}
    \label{alg:methods:rnn-search}
    \begin{tiny}
    \begin{algorithmic}
        \Require $root$, the root cluster
        \Require $q$, a query
        \Require $\rho$, a search radius

        \State $S$ $\Leftarrow$ tree-search($root$, $q$, $\rho$)
        \State $H$ $\Leftarrow$ leaf-search($S$, $q$, $\rho$)
        \State \textbf{return} $H$
    \end{algorithmic}
    \end{tiny}
\end{algorithm}

\begin{gather}
    \mathcal{O}
    \Bigg(
        \underbrace{
            \log~\overbrace{\mathcal{N}_{\hat{r}}(X)}^{\textrm{metric entropy}}
        }_{\textrm{tree-search}}
        \ + \
        \underbrace{
            \overbrace{ \big| B_X(q, \rho) \big|}^{\textrm{output size}}
            \overbrace{ \left( \frac{\rho + 2 \cdot \hat{r}}{ \rho} \right) ^ d}^{\textrm{scaling factor}}
        }_{\textrm{leaf-search}}
    \Bigg)
    \label{eq:methods:rnn-search-complexity}
\end{gather}
where $\hat{r}$ is the \textit{mean} radius of leaf clusters, $\mathcal{N}_{\hat{r}}(X)$ is the metric entropy at that radius, $B_X(q, \rho)$ is a ball of radius $\rho$ around the query $q$, and $d$ is the LFD around the query at the length scale of $\rho$ and $\rho + 2 \cdot \hat{r}$.

From Algorithm~\ref{alg:methods:rnn-search}, we have that $H = B_X(q, \rho)$, and so $\rho$-NN search performance scales linearly with the size of the output set and exponentially with the LFD around the query.
While one might worry that the exponential scaling factor would dominate, the manifold hypothesis suggests that the LFD of real-world datasets is typically very low, and so this algorithm actually scales sub-linearly with the cardinality of the dataset.

@@

@@

In these algorithms, we use $H$, for \textit{hits}, to refer to the data structure which stores the closest points to the query found so far, and $Q$ to refer to the data structure which stores the clusters and points which are still in contention for being one of the $k$ nearest neighbors.

@@

In this algorithm, we perform $\rho$-NN search starting with a small search radius, and repeatedly increasing the radius until $|H| \geq k$.

Let the search radius $r$ be equal to the radius of the root cluster divided by the cardinality of the dataset.
We perform tree-search with radius $r$.
If no clusters are found, then we double $r$ and perform tree-search again, repeating until we find at least one cluster.
Let $Q$ be the set of clusters returned by the first tree-search which returns at least one cluster.

>>>>

\begin{equation}
    \text{LFD}(q, r_1, r_2) = \frac{\text{log} \left( \frac{|B(q, r_1)|}{|B(q, r_2)|} \right) }{\text{log} \left( \frac{r_1}{r_2} \right) }
    \label{eq:methods:lfd-original}
\end{equation}

where $B(q, r)$ is the set of points contained in the metric ball of radius $r$ centered at a point $q$ in the dataset $\textbf{X}$.

We use a simplified version of this definition by calculating the LFD at a length scale where $r_1 = 2 \cdot r_2$.

\begin{equation}
    \text{LFD}(q, r) = \text{log}_2 \left( \frac{|B(q, r)|}{|B(q, \frac{r}{2})|} \right)
    \label{eq:methods:lfd-half}
\end{equation}

Intuitively, LFD measures the rate of change in the number of points in a ball of radius $r$ around a point $q$ as $r$ increases. When the vast majority of points in the dataset have low ($\ll D$) LFD, we can simply say that the dataset has low LFD.
We stress that this concept differs from the \textit{embedding dimension} of a dataset.
To illustrate the difference, consider the SILVA 18S rRNA dataset which contains genomic sequences with unaligned lengths of up to 3,718 base pairs and aligned length of 50,000 base pairs.
Hence, the \textit{embedding dimension} of this dataset is at least 3,718 and at most 50,000.
However, physical constraints (namely, biological evolution and biochemistry) constrain the data to a lower-dimensional manifold within this space.
LFD is an approximation of the dimensionality of that lower-dimensional manifold in the ``vicinity'' of a given point.
Figure~\ref{fig:results:lfd-plots} illustrates this concept on a variety of datasets, showing how real datasets uphold the manifold hypothesis.
For real-world datasets, we expect the LFD to be locally uniform, i.e., when $r$ is small, but potentially highly variable at larger length scales, i.e., when $r$ is large.

@@

We define a \textit{cluster} as a set of points with a \textit{center} and a \textit{radius}.
The center is the geometric median of the points in the cluster, i.e., it is the point which minimizes the sum of distances to all other points in the cluster.
In cases where the cardinality of the cluster is large, we take a random subsample of $\sqrt{|C|}$ points and compute the geometric median of that subsample~\cite{ishaq2019clustered}.
The center, therefore, is one of the points in the cluster and is used as a representative of the cluster.
The radius is the maximum distance from the center to any point in the cluster.
Each non-leaf cluster has two child clusters in much the same way that a node in a binary tree has two child nodes.
Note that clusters can have overlapping volumes and, in such cases, points in the overlapping volume are assigned to exactly one of the overlapping clusters.
As a consequence, a cluster can be a proper subset of the metric ball at the same center and radius, i.e. $C(c, r) \subset B(c, r)$.

Hereafter, when we refer to the LFD of a cluster, it is estimated at the length scale of the cluster radius and half that radius, i.e. using Equation~\ref{eq:methods:lfd-half}.
We also only use points that are in $C(c, r)$ instead of all those in $B(c, r)$.

The \textit{metric entropy} $\mathcal{N}_{r}(X)$ for some radius $r$ is the minimum number of clusters of a uniform radius $r$ needed to cover the data~\cite{yu2015entropy} for a flat clustering with clusters of radius $r$.
In this paper, where the clustering is hierarchical rather than flat, we define the metric entropy $\mathcal{N}_{\hat{r}}(X)$ as the number of leaf clusters in the tree where $\hat{r}$ is the mean radius of all leaf clusters.

@@

We start by performing a divisive hierarchical clustering on the dataset using CLAM.
The procedure is almost identical to that outlined in CHESS~\cite{ishaq2019clustered}, but with better selection of poles for partitioning (see Algorithm~\ref{alg:methods:partition}).
After building the tree, we also perform a depth-first reordering of the dataset (see Section~\ref{sec:methods:clustering:depth-first-reordering}).

For a cluster $C$ with $|C|$ points, we begin by taking a random sample of $\sqrt{|C|}$ of its points, and computing pairwise distances between all points in this sample.
Using these distances, we compute the \textit{geometric median} of this sample; in other words, we find the point which minimizes the sum of distances to all other points in the sample.
This geometric median is the center of $C$.

The \textit{radius} of $C$ is the maximum distance from the \textit{center} to any other point in $C$.
The point which is responsible for that radius (i.e., the furthest point from the center) is designated the \textit{left pole} and the point which is furthest from the left pole is designated the \textit{right pole}.
We then partition the cluster into a \textit{left child} and a \textit{right child}, where the left child contains all points in the cluster which are closer to the left pole than to the right pole, and the right child contains all points in the cluster which are closer to the right pole than to the left pole.
Without loss of generality, we assign to the left child those points which are equidistant from the two poles.
Starting from a root cluster containing the entire dataset, we repeat this procedure until each leaf contains only one datum, or we meet some other user-specified stopping criteria, for example, minimum cluster radius, minimum cluster cardinality, maximum tree depth, etc.
This process is described in Algorithm \ref{alg:methods:partition}.
During the partitioning process, we also compute (and cache) the LFD of each cluster using Equation~\ref{eq:methods:lfd-half}.

\begin{algorithm} % enter the algorithm environment
    \caption{Partition($C$, $criteria$)} % give the algorithm a caption
    \label{alg:methods:partition} % and a label for \ref{} commands later in the document
    \begin{algorithmic}[0] % enter the algorithmic environment
        \Require $C$, a cluster
        \Require $criteria$, user-specified continuation criteria

@@

@@

This complexity analysis ignores the cost of checking the continuation criteria.
While a user could specify criteria of any complexity (our implementation allows for this), for this paper we continue clustering until each cluster is a singleton, i.e., it contains only one point or duplicates of the same point and has a radius of zero.
This criterion costs $\mathcal{O}(1)$ to check per cluster, and so does it not affect the overall complexity of the algorithm as used in this paper.

We also note here that this $\mathcal{O}(n \log n)$ complexity assumes that the tree is balanced.
This can realistically only happen when the data are uniformly distributed in the embedding space.
This represents the worst-case scenario for subsequent search algorithms because the LFD of the dataset will approach or equal the embedding dimension.
For real-world datasets with low LFD, we expect the tree to be highly unbalanced, reflecting the topological structure of the data.
When building trees for such datasets, their depths would be larger than $\log n$, which would seem to increase the cost of building
However, because of the large numbers of leaf clusters at shallow depths, each subsequent level of the tree would also have fewer points, and so the cost of building each subsequent level would be lower.
Any analysis of unbalanced trees would be highly dependent on the specific dataset, and so we do not provide one here.
In any case, the tree is built only once, so this cost is amortized over all search queries and other applications.

% TODO: Move most of this to the supplement
% \subsection{\texorpdfstring{$\rho$}{p}-Nearest Neighbors Search}
% \label{sec:methods:rnn-search}

% We conduct $\rho$-NN search as described in~\cite{ishaq2019clustered}, but with the following improvement:
% when a cluster overlaps with the query ball, instead of always proceeding to search both of its children, we proceed only with those children which might contain points in the query ball.

% \begin{figure}[ht!]
%     \centering
%     \includegraphics[scale=0.75]{images/geometry/overlapping-children-3.pdf}
%     \caption{The geometry of a query ball overlapping with a cluster and either one or both of its children. Here, $l$ is the left pole, $r$ is the right pole, and $q$ is the query. Other points and distances are described in the text.}
%     \label{fig:methods:overlapping-children}
%     \caption{CAKES uses geometric properties of clusters.}
% \end{figure}

% To determine whether both children can contain points in the query ball, we consider Figure~\ref{fig:methods:overlapping-children}.
% Here, we overload the notation for $\overline{x y}$ to refer both to the line segment joining points $x$ and $y$ as well as to the length of that line segment.

% Let $q$ denote the query, $\rho$ denote the search radius, and $l$ and $r$ denote the cluster's left and right poles respectively (see Section~\ref{sec:methods:clustering:building-the-tree}).
% Without loss of generality, we assume that $\overline{q r \vphantom{l}} \leq \overline{q l}$.
% Now let $q'$ be the projection of $q$ onto $\overline{l r}$, $m$ be the midpoint of $\overline{l r}$, and $d$ be the distance from $q'$ to $m$.
% As a consequence of how we assign a point in the parent cluster to the left child in Algorithm~\ref{alg:methods:partition}, if $\rho < d$, then the left child cannot contain points inside the query ball.
% In such a case we proceed to search only the right child.
% Otherwise, we proceed with both children.

% To check whether $d \leq \rho$, we note that $d = \overline{m q' \vphantom{l}} = \overline{m r \vphantom{l}} - \overline{q' r \vphantom{l}} = \frac{\overline{l r}}{2} - \overline{q' r \vphantom{l}}$.
% Let $\theta$ denote $\angle l r q$, as shown in Figure~\ref{fig:methods:overlapping-children}.
% By the Law of Cosines on $\triangle l r q$, we have that $\text{cos}(\theta) = \tfrac{\overline{l r}^2 + \ \overline{q r \vphantom{l}}^2 - \ \overline{q l}^2}{2 \cdot \overline{l r} \cdot \overline{q r \vphantom{l}}}$.
% Since $\triangle r q q'$ is a right triangle, we also have that $\text{cos}(\theta) = \tfrac{\overline{q' r \vphantom{l}}}{\overline{q r \vphantom{l}}}$.
% Combining the previous two equations and solving for $\overline{q' r \vphantom{l}}$, we have that $\overline{q' r \vphantom{l}} = \tfrac{\overline{q r \vphantom{l}}^2 + \ \overline{l r}^2 - \ \overline{q l}^2}{2 \cdot \overline{l r}}$.
% Substituting for $\overline{q' r \vphantom{l}}$ in the equation for $d$, we have that $d = \tfrac{\overline{l r}}{2} - \tfrac{\overline{q r \vphantom{l}}^2 + \ \overline{l r}^2 - \ \overline{q l}^2}{2 \cdot \overline{l r}} = \tfrac{\overline{q l}^2 - \overline{q r \vphantom{l}}^2}{2 \cdot \overline{l r}}$.

% Thus, $d \leq \rho \iff (\overline{q l} + \overline{q r \vphantom{l}})(\overline{q l} - \overline{q r \vphantom{l}}) \leq 2 \cdot \overline{l r} \cdot \rho$. Note, in particular, that this only requires distances between actual points from the dataset, and so it can be used with any distance function, even when $q'$ and $m$ are not actual points or cannot be imputed from the data.

% To perform $\rho$-NN search, we first perform a coarse \textit{tree-search}, as outlined in Algorithm~\ref{alg:methods:rnn-search:tree-search}, to find the leaf clusters which overlap with the query ball or any clusters which lie entirely within the query ball.
% Then, for all such clusters, we perform a finer-grained \textit{leaf-search}, as outlined in Algorithm~\ref{alg:methods:rnn-search:leaf-search}, to find all points which are no more than a distance $\rho$ from the query.
% The asymptotic complexity of $\rho$-NN is the same as in~\cite{ishaq2019clustered} and shown in Equation~\ref{eq:methods:rnn-search-complexity}.

% \begin{gather}
%     \mathcal{O}
%     \Bigg(
%         \underbrace{
%             \log~\overbrace{\mathcal{N}_{\hat{r}}(X)}^{\textrm{metric entropy}}
%         }_{\textrm{tree-search}}
%         \ + \
%         \underbrace{
%             \overbrace{ \big| B_X(q, \rho) \big|}^{\textrm{output size}}
%             \overbrace{ \left( \frac{\rho + 2 \cdot \hat{r}}{ \rho} \right) ^ d}^{\textrm{scaling factor}}
%         }_{\textrm{leaf-search}}
%     \Bigg)
%     \label{eq:methods:rnn-search-complexity}
% \end{gather}
% where $\hat{r}$ is the \textit{mean} radius of leaf clusters, $\mathcal{N}_{\hat{r}}(X)$ is the metric entropy at that radius, $B_X(q, \rho)$ is a ball of radius $\rho$ around the query $q$, and $d$ is the LFD around the query at the length scale of $\rho$ and $\rho + 2 \cdot \hat{r}$.

@@

Given a dataset $\textbf{X}$, a distance function $f$ and the root cluster of the tree built using Algorithm~\ref{alg:methods:partition}, we can now formally pose the $k$-NN search problem.

The $k$-NN search problem, given a user specified integer $k$ and query $q$, aims to find the $k$ closest points to $q$ in $\textbf{X}$.
In other words, $k$-NN search aims to find the set $H$ such that $|H| = k$ and $H = B_X(q, \rho_k)$ where $\rho_k = \max \left\{ f(q, p) \ \forall \ p \in H \right\}$ is the distance from $q$ to the $k^{th}$ nearest neighbor in $\textbf{X}$.

@@

In these algorithms, we use $H$, for \textit{hits}, to refer to the data structure which stores the closest points to the query found so far, and $Q$ to refer to the data structure which stores the clusters and points which are still in contention for being one of the $k$ nearest neighbors.
The two Sieve algorithms also use some terminology defined and illustrated in Figure~\ref{fig:methods:deltas}.

\begin{figure}
    \centering
    \includegraphics[scale=0.65]{images/geometry/deltas.pdf}
    \caption{
        {\color{blue}$\delta$}, {\color{red}$\delta^{+}$}, and {\color{green}$\delta^{-}$} for a cluster $C$ and a query $q$.
        ${\color{blue}\delta} = f(q, c)$ is the distance from the query to the cluster center $c$.
        ${\color{red}\delta^{+}} = \delta + r$ is the distance from the query to the theoretically farthest point in $C$.
        ${\color{green}\delta^{-}} = \text{max}(0, \delta - r)$ is the distance from the query to the theoretically closest point in $C$.
    }
    \label{fig:methods:deltas}
\end{figure}%

@@

This algorithm relies on the \textit{tree-search} and \textit{leaf-search} algorithms described in~\cite{ishaq2019clustered}.
We reproduce the algorithms here for completeness.

\begin{minipage}{.425\textwidth}
    \begin{algorithm}[H]\small
        \caption{tree-search($C$, $q$, $r$)}
        \label{alg:methods:rnn-search:tree-search}
        \begin{algorithmic}
            \Require $C$, a cluster
            \Require $q$, a query
            \Require $r$, a search radius

            \If{$C$ is a leaf \textbf{or} $\delta^+_C \leq r$}
                \State \textbf{return} $\{C\}$
            \Else
                \State $[l, r]$ $\Leftarrow$ \textit{pole}s of $C$
                \State $[L, R]$ $\Leftarrow$ \textit{children} of $C$
                \State \textbf{return} tree-search($L, q, r$)

                \ \ \ \ \ \ \ \ \ $\cup$ tree-search($R, q, r$)
            \EndIf

        \end{algorithmic}
    \end{algorithm}
\end{minipage}
\hfill
\begin{minipage}{.45\textwidth}
    \begin{algorithm}[H]\small
        \caption{leaf-search($Q$, $q$, $r$)}
        \label{alg:methods:rnn-search:leaf-search}
        \begin{algorithmic}
            \Require $Q$, a set of clusters
            \Require $q$, a query
            \Require $r$, a search radius

            \State $H \Leftarrow \emptyset$, a set of 2-tuples of points and their distances to the query

            \For{$C \in Q$}
                \If{$\delta^+_C \leq r$}
                    \State $H$ $\Leftarrow$ $H \cup \{C\}$
                \Else
                    \For{$p \in C$}
                        \If{$f(p, q) \leq r$}
                            \State $H$ $\Leftarrow$ $H \cup \{(p, f(p, q))\}$
                        \EndIf
                    \EndFor
                \EndIf
            \EndFor

            \State \textbf{return} $H$
        \end{algorithmic}
    \end{algorithm}
\end{minipage}

For Repeated $\rho$-NN search, we begin by performing \textit{tree-search} (Algorithm~\ref{alg:methods:rnn-search:tree-search}) with a search radius $r$ equal to the radius of the root cluster divided by the cardinality of the dataset.
If no clusters are found, then we double $r$ and perform tree-search again, repeating until we find at least one cluster.
